---
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=FALSE)
options(repos = c(CRAN = "https://cloud.r-project.org/"))

#Install required libraries
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("SmartEDA")
install.packages("plotly")
install.packages("ggthemes")
install.packages("dplyr")
install.packages("corrplot")
install.packages("ggcorrplot")
install.packages("gridExtra")
install.packages("flexdashboard")
install.packages("shiny")


#Load required libraries
library(flexdashboard)
library(tidyverse)
library(dplyr)
library(SmartEDA)
library(plotly)
library(ggplot2)
library(GGally)
library(leaflet)
library(ggthemes)
library(corrplot)
library(ggcorrplot)
library(gridExtra)
library(grid)
library(shiny)
theme_set(theme_classic())
```


```{r, results='hide'}
#Import Dataset
data <- read.csv("~/Downloads/37-00049_UOF-P_2016_prepped.csv")

# View the structure of the dataset
str(data)

# Summary of the dataset
summary(data)

#Get the shape of the dataset
dim(data)
```


```{r}
#converting to numerical
data$OFFICER_YEARS_ON_FORCE <- as.numeric(data$OFFICER_YEARS_ON_FORCE)
data$STREET_NUMBER <- as.numeric(data$STREET_NUMBER)
data$SECTOR <- as.numeric(data$SECTOR)
data$OFFICER_ID <- as.numeric(data$OFFICER_ID)
data$SUBJECT_ID <- as.numeric(data$SUBJECT_ID)
data$REPORTING_AREA <- as.numeric(data$REPORTING_AREA)
data$BEAT <- as.numeric(data$BEAT)
data$LOCATION_LATITUDE <- as.numeric(data$LOCATION_LATITUDE)
data$LOCATION_LONGITUDE <- as.numeric(data$LOCATION_LONGITUDE)

# Convert INCIDENT_DATE to Date format
data$INCIDENT_DATE <- as.Date(data$INCIDENT_DATE, format = "%m/%d/%y")

# convert all characters to factors
char_cols <- sapply(data, is.character)
data[char_cols] <- lapply(data[char_cols], as.factor)

# Delete the first row
data <- data[-1, ]

# Delete empty columns
data <- data[, -c(37,38,39,40,41,42,43,44,45)]
```


```{r}
# Select only numeric columns
numeric_columns <- data %>%
  select_if(is.numeric)

# Compute the correlation matrix
correlation_matrix <- cor(numeric_columns, use = "complete.obs")

# Calculate the p-value matrix
p_value_matrix <- cor_pmat(numeric_columns)

# Create ggcorrplot with p-values and hierarchical clustering order
G1 <- ggcorrplot(correlation_matrix, p.mat = p_value_matrix, hc.order = TRUE, type = "lower")

# Add title to ggcorrplot
G1 <- G1 + ggtitle("Correlation Matrix with P-values")

# Display ggcorrplot
G1

# Create corrplot with circle method, p-values, significance level, and blank insignificance
G2 <- corrplot(correlation_matrix, method = "circle", p.mat = p_value_matrix, sig.level = 0.05, insig = "blank")
```


# Two-way table of officer race and subject race
two_way_table <- table(data$OFFICER_RACE, data$SUBJECT_RACE)
print(two_way_table)
```


```{r}
#Count of use of force incidents by officer race
data %>%
  ggplot(aes(x = OFFICER_RACE, fill = OFFICER_RACE)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(
    title = "Use of Force Incidents by Officer Race",
    x = "Officer Race",
    y = "Count"
  )
```


```{r}
#Count of use of force incidents by subject race
data %>%
  ggplot(aes(x = SUBJECT_RACE, fill = SUBJECT_RACE)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set3") +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(
    title = "Use of Force Incidents by Subject Race",
    x = "Subject Race",
    y = "Count"
  )

```

```{r}
#Count of use of force incidents by officer gender
data %>%
  ggplot(aes(x = OFFICER_GENDER, fill = OFFICER_GENDER)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(
    title = "Use of Force Incidents by Officer Gender",
    x = "Officer Gender",
    y = "Count"
  )

```


```{r}
#Count of use of force incidents by subject gender
data %>%
  ggplot(aes(x = SUBJECT_GENDER, fill = SUBJECT_GENDER)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(
    title = "Use of Force Incidents by Subject Gender",
    x = "Subject Gender",
    y = "Count"
  )

```

### Geographical View


```{r, warning=FALSE}
#Geographical distribution of use of force incidents (using latitude and longitude)
data %>%
  ggplot(aes(x = LOCATION_LONGITUDE, y = LOCATION_LATITUDE)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Geographical Distribution of Use of Force Incidents",
       x = "Longitude",
       y = "Latitude")

```


```{r}
# Filter out rows with missing latitude and longitude
map_data <- data %>%
  filter(!is.na(LOCATION_LATITUDE) & !is.na(LOCATION_LONGITUDE))

# Convert latitude and longitude to numeric
map_data$LOCATION_LATITUDE <- as.numeric(map_data$LOCATION_LATITUDE)
map_data$LOCATION_LONGITUDE <- as.numeric(map_data$LOCATION_LONGITUDE)

# Create a Leaflet map
leaflet(map_data) %>%
  addTiles() %>%
  addCircleMarkers(lng = ~LOCATION_LONGITUDE, lat = ~LOCATION_LATITUDE,
                   radius = 2, stroke = FALSE, fillOpacity = 0.5) %>%
  addControl(html = "<div style='background-color: rgba(255, 255, 255, 0.8); padding: 5px;'><strong>Geolocated Data Points Visualization with Leaflet</strong></div>", position = "topright")
```


```{r}
# Combine subject race and officer race into a single data frame
combined_data <- data %>%
  gather(key = "Race_Type", value = "Race", SUBJECT_RACE, OFFICER_RACE)

# Create the combined bar chart
combined_bar_chart <- ggplot(combined_data, aes(x = DIVISION, y = ..count.., fill = Race)) +
  geom_bar(position = "dodge", width = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_text(size  = 9,
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        plot.title = element_text(size = 14, hjust = 0.5)) +
  xlab("Division") +
  ylab("Count") +
  ggtitle("Distribution of Subject and Officer Race Across Divisions") +
  guides(fill = guide_legend(title = "Race")) +
  facet_wrap(~ Race_Type, ncol = 1, scales = "free_y")

# Make the bar chart interactive
interactive_combined_bar_chart <- ggplotly(combined_bar_chart)

# Print the interactive combined bar chart
interactive_combined_bar_chart
```

```{r}
# Time series plot for daily number of incidents
p <- data %>%
  group_by(INCIDENT_DATE) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = INCIDENT_DATE, y = count)) +
  geom_line(color = "steelblue", size = 0.6) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  labs(title = "Daily Number of Incidents",
       x = "Date",
       y = "Number of Incidents")

# Make the plot interactive
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot

```


```{r}
# Smoothing using geom_smooth()
data %>%
  group_by(INCIDENT_DATE) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = INCIDENT_DATE, y = count)) +
  geom_point() +
  geom_smooth(method = "loess", span = 0.3) +
  labs(title = "Daily Number of Incidents with Trend",
       x = "Date",
       y = "Number of Incidents")

```

```{r}
# Pie chart of Officer injury
pie_data <- data.frame(table(data$OFFICER_INJURY))
pie_data$Var1 <- as.factor(pie_data$Var1)

# Calculate the percentages and add them to the data frame
total_freq <- sum(pie_data$Freq)
pie_data$Percentage <- round((pie_data$Freq / total_freq) * 100, 1)

# Create the pie chart with percentage labels
ggplot(pie_data, aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  geom_text(aes(y = Freq / 2 + c(0, cumsum(Freq)[-length(Freq)]), label = paste0(Percentage, "%")), size = 4) +
  theme_minimal() +
  labs(title = "Officer Injury (Pie Chart)") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank())

# Pie chart of subject injury
pie_data <- data.frame(table(data$SUBJECT_INJURY))
pie_data$Var1 <- as.factor(pie_data$Var1)

# Calculate the percentages and add them to the data frame
total_freq <- sum(pie_data$Freq)
pie_data$Percentage <- round((pie_data$Freq / total_freq) * 100, 1)

# Create the pie chart with percentage labels
ggplot(pie_data, aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  geom_text(aes(y = Freq / 2 + c(0, cumsum(Freq)[-length(Freq)]), label = paste0(Percentage, "%")), size = 4) +
  theme_minimal() +
  labs(title = "Subject Injury (Pie Chart)") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank())
```

```{r}
top_n <- 10 # Set the number of top injuries to display

# Process officer injury data
officer_injury_data <- data %>%
  filter(!is.na(OFFICER_INJURY_TYPE)) %>%
  group_by(OFFICER_INJURY_TYPE) %>%
  summarize(count = n()) %>%
  mutate(type = "Officer") %>%
  rename(injury = OFFICER_INJURY_TYPE) %>%
  top_n(n = top_n, wt = count)

# Process subject injury data
subject_injury_data <- data %>%
  filter(!is.na(SUBJECT_INJURY_TYPE)) %>%
  group_by(SUBJECT_INJURY_TYPE) %>%
  summarize(count = n()) %>%
  mutate(type = "Subject") %>%
  rename(injury = SUBJECT_INJURY_TYPE) %>%
  top_n(n = top_n, wt = count)

# Combine officer and subject injury data
injury_data <- rbind(officer_injury_data, subject_injury_data)

# Create the bar chart
bar_chart <- ggplot(injury_data, aes(x = injury, y = count, fill = type, color = type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  labs(title = "Comparison of Top Officer and Subject Injury Type",
       x = "Injury",
       y = "Count",
       fill = "Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert ggplot to plotly object
interactive_bar_chart <- ggplotly(bar_chart)

# Display the interactive bar chart
interactive_bar_chart

```


```{r}
# Dot plot of use of force incidents by officer race with enhanced aesthetics and gridlines
p <- ggplot(data, aes(x = OFFICER_RACE , y = after_stat(count), group = OFFICER_RACE, color = OFFICER_RACE)) +
  geom_point(stat = "count", size = 3, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Interactive Dot Plot",
    x = "Officer Race",
    y = "Count"
  )

# Make the plot interactive
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot
```



```{r}
# Histogram of officer years on force with enhanced aesthetics
ggplot(data, aes(x = OFFICER_YEARS_ON_FORCE)) +
  geom_histogram(binwidth = 1, fill = "#56B4E9", color = "black", alpha = 0.8) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Histogram of Officer Years on Force",
    x = "Years on Force",
    y = "Frequency"
  )
```


```{r}
# Box plot of officer years on force by officer race with enhanced aesthetics
p <- ggplot(data, aes(x = OFFICER_RACE, y = OFFICER_YEARS_ON_FORCE, fill = OFFICER_RACE)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Interactive Box Plot",
    x = "Officer Race",
    y = "Years on Force"
  )

# Make the plot interactive
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot

```


```{r}
# Violin plot of officer years on force by officer race
p <- ggplot(data, aes(x = OFFICER_RACE, y = OFFICER_YEARS_ON_FORCE, fill = OFFICER_RACE)) +
  geom_violin(alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(title = "Interactive Violin Plot",
       x = "Officer Race",
       y = "Years on Force")

# Make the plot interactive
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot

```


```{r}
# Create a scatter plot using the policing dataset
p <- ggplot(data, aes(x = INCIDENT_REASON, y = OFFICER_YEARS_ON_FORCE, color = OFFICER_RACE)) +
  geom_jitter(width = 0.2, height = 0.3, alpha = 1.7) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Scatter plot of Officer Years on Force vs. Incident Reason",
       x = "Incident Reason",
       y = "Officer Years on Force",
       color = "Officer Race")

# Make the plot interactive
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot
```

